Generating Random Logic Programs Using Constraint Programming

### 2020-01-12 paper/paper.tex
Some writing

TODO: define all the relevant terminology from logic. If a predicate has arity
$n$ and it is as of yet undecided what $n$ terms will fill those spots, we will
say that the predicate has $n$ \emph{gaps}.

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (or) at (0, 0) {$\lor$};
    \node[draw,circle,gray,text=black] (not) at (-1, -1) {$\neg$};
    \node (P) at (-1, -2) {$\mathsf{P}(X)$};
    \node[draw,circle,gray,text=black] (and) at (1, -1) {$\land$};
    \node (Q) at (0, -2) {$\mathsf{Q}(X)$};
    \node (R) at (2, -2) {$\mathsf{P}(X)$};
    \draw[gray] (or) -- (not);
    \draw[gray] (or) -- (and);
    \draw[gray] (not) -- (P);
    \draw[gray] (and) -- (Q);
    \draw[gray] (and) -- (R);
  \end{tikzpicture}
  \caption{A tree representation of the formula from \cref{example:formula}}
  \label{fig:example_tree}
\end{figure}

@@

\begin{definition}
  Let $\mathscr{P}$ be a probabilistic logic program. Its \emph{predicate
    dependency graph} is a directed graph $G_{\mathscr{P}} = (V, E)$ with the
  set of nodes $V$ consisting of all predicates in $\mathscr{P}$. We add an edge
  from predicate $\mathsf{P}$ to predicate $\mathsf{Q}$ if there is a clause in
  $\mathscr{P}$ with $\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the
  body.
\end{definition}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw] (father) at (0, 0.5) {$\mathsf{father}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (parent) at (2, 0) {$\mathsf{parent}$};
    \node[draw] (sibling) at (4, 0) {$\mathsf{sibling}$};
    \draw[-{Stealth[scale=1.5]}] (father) -- (parent);
    \draw[-{Stealth[scale=1.5]}] (mother) -- (parent);
    \draw[-{Stealth[scale=1.5]}] (parent) -- (sibling);
  \end{tikzpicture}
  \caption{The predicate dependency graph of the program in \cref{ex:program}}
  \label{fig:predicate_dependencies}
\end{figure}

\begin{example} \label{ex:program}
  Consider the following program:
  \begin{align*}
    \mathsf{sibling}(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z),\\
    \mathsf{parent}(X, Y) &\gets \mathsf{father}(X, Y) \lor \mathsf{mother}(X, Y).
  \end{align*}
  Its predicate dependency graph is in \cref{fig:predicate_dependencies}. We can
  then list the dependencies of each predicate:
  \begin{alignat*}{3}
    D_{\mathsf{father}} &= \{ \mathsf{father} \}, \quad && D_{\mathsf{parent}} &&= \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent} \}, \\
    D_{\mathsf{mother}} &= \{ \mathsf{mother} \}, \quad && D_{\mathsf{sibling}} &&= \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent}, \mathsf{sibling} \}.
  \end{alignat*}
  Hence, the only pair of independent predicates in this program is
  $\mathsf{father}$ and $\mathsf{mother}$.
\end{example}

% TODO: continue writing from here
\begin{definition}[Adjacency matrix representation]
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$
  defined by
  \begin{align*}
    A_{i,j} = 0 \iff &\nexists k: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{name} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{definition}

A dependency is an algebraic data type that is either determined (in which case
it holds only the index of the predicate) or undetermined (in which case it also
holds the indices of the source and target vertices, corresponding to the edge
responsible for making the dependency undetermined).

Propagation for independence:
\begin{itemize}
\item Two types of dependencies: determined and
  one-undetermined-edge-away-from-being-determined.
\item Look up the dependencies of both predicates. For each pair of
  matching dependencies:
  \begin{itemize}
  \item If both are determined, fail.
  \item If one is determined, the selected edge of the other must not
    exist.
  \end{itemize}
\end{itemize}

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$} \times
    \getDependencies{$p_2$}$ s.t. $d_1.\predicate = d_2.\predicate$}{
    \If{$d_1$.\isDetermined{} {\bf and} $d_2$.\isDetermined{}}{
      \fail{}\;
    }
    \uIf{$d_1$.\isDetermined{}}{
      $\mathbf{A}[d_2.\source][d_2.\target]$.\removeValue{$1$}\;
    }
    \ElseIf{$d_2$.\isDetermined{}}{
      $\mathbf{A}[d_1.\source][d_1.\target]$.\removeValue{$1$}\;
    }
  }
  \caption{Propagation}
\end{algorithm}

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$} \times
  \getDependencies{$p_2$} \mid d_1.\predicate = d_2.\predicate \}$\;
  \If{$\{ (d_1, d_2) \in D \mid d_1.\isDetermined{}, d_2.\isDetermined{} \} \ne
    \emptyset$}{
    \Return{FALSE}\;
  }
  \If{$D = \emptyset$}{
    \Return{TRUE}\;
  }
  \Return{UNDEFINED}\;
  \caption{Entailment}
\end{algorithm}

\begin{algorithm}[p]
  \SetKwData{edgeExists}{edgeExists}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{an $n \times n$ adjacency matrix $\mathbf{A}$}
  \Fn{\getDependencies{$p$}} {
    $D \gets \{ p \}$\;
    \Repeat{$D' = D$}{
      $D' \gets D$\;
      \For{$d \in D$}{
        \For{$i \gets 1$ \KwTo $n$}{
          $\edgeExists \gets \mathbf{A}[i][d.\predicate] = \{ 1 \}$\;
          \uIf{$\edgeExists$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ i \}$\;
          }
          \uElseIf{$\edgeExists$ {\bf and not} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, d.\source, d.\target) \}$\;
          }
          \ElseIf{$|\mathbf{A}[i][d.\predicate]| > 1$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, i, d.\predicate) \}$\;
          }
        }
      }
    }
    \Return{$D$}\;
  }
  \caption{Computing the dependencies of a predicate}
\end{algorithm}

\section{Entailment Checking for Negative/All Cycles}

\begin{enumerate}
\item Let $C$ be a set of clauses such that their bodies and predicates in their
  heads are fully determined.
\item If $C = \emptyset$, return UNDEFINED.
\item Construct an adjacency list representation of a graph where vertices
  represent predicates. Each edge is either \emph{positive} or \emph{negative}.
  There is an edge from $p$ to $q$ if $q$ appears in the body of a predicate
  with $p$ as its head. The edge is negative if, when traversing the tree to
  reach some instance of $q$, we pass through a $\neg$ node. Otherwise, it's
  positive.
\item Run a modified cycle detection algorithm that detects all cycles that have
  at least one negative edge.
\item If we found a cycle, return FALSE.
\item If $C$ encompasses all clauses, return TRUE.
\item Return UNDEFINED.
\end{enumerate}

>>>>

TODO: define all the relevant terminology from logic and constraint programming.

If a predicate has arity $n$ and it is as of yet undecided what $n$ terms will
fill those spots, we will say that the predicate has $n$ \emph{gaps}. We say
that a constraint variable is \emph{(fully) determined} if its domain (at the
given moment in the execution) has exactly one value.

A \emph{(logic) program} is a multiset of clauses. Given a program
$\mathscr{P}$, a \emph{subprogram} $\mathscr{R}$ of $\mathscr{P}$ is a subset of
the clauses of $\mathscr{P}$ and is denoted by $\mathscr{R} \subseteq
\mathscr{P}$.

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (or) at (-0.375, 0) {$\lor$};
    \node[draw,circle,gray,text=black] (not) at (-1.5, -1) {$\neg$};
    \node[draw,circle,gray,text=black] (and) at (0.75, -1) {$\land$};
    \node[draw,gray,text=black] (P) at (-1.5, -2) {$\mathsf{P}(X)$};
    \node[draw,gray,text=black] (Q) at (0, -2) {$\mathsf{Q}(X)$};
    \node[draw,gray,text=black] (R) at (1.5, -2) {$\mathsf{P}(X)$};
    \draw[gray] (or) -- (not);
    \draw[gray] (or) -- (and);
    \draw[gray] (not) -- (P);
    \draw[gray] (and) -- (Q);
    \draw[gray] (and) -- (R);
  \end{tikzpicture}
  \caption{A tree representation of the formula from \cref{example:formula}}
  \label{fig:example_tree}
\end{figure}

@@

\begin{definition}
  Let $\mathscr{P}$ be a probabilistic logic program. Its \emph{predicate
    dependency graph} is a directed graph $G_{\mathscr{P}} = (V, E)$ with the
  set of nodes $V$ consisting of all predicates in $\mathscr{P}$. For any two
  different predicates $\mathsf{P}$ and $\mathsf{Q}$, we add an edge from
  $\mathsf{P}$ to $\mathsf{Q}$ if there is a clause in $\mathscr{P}$ with
  $\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say that
  the edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as the
  head and at least one instance of $\mathsf{P}$ at the body such that the path
  from the root to the $\mathsf{P}$ node in the tree representation of the
  clause passes through at least one negation node. Otherwise it is
  \emph{positive}. We say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a
  \emph{negative cycle} if $G_{\mathscr{P}}$ has a cycle with at least one
  negative edge.
\end{definition}

@@

\begin{figure}[t]
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
      \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
      \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
      \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
      \draw[gray] (and) -- (parent);
      \draw[gray] (and) -- (not);
      \draw[gray] (not) -- (mother);
    \end{tikzpicture}
    \caption{The tree representation of the body of Clause~\eqref{eq:example_clause}.}
    \label{fig:example_tree2}
  \end{minipage}
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
      \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
      \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
      \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
      \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
    \end{tikzpicture}
    \caption{The predicate dependency graph of the program in \cref{ex:program}.
      Positive edges are labeled with `$+$', and negative edges with `$-$'.}
    \label{fig:predicate_dependencies}
  \end{minipage}
\end{figure}

% TODO: example of edge positivity/negativity
\begin{example} \label{ex:program}
  Consider the following program:
  \begin{align}
    \mathsf{sibling}(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z), \nonumber \\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y). \label{eq:example_clause}
  \end{align}
  Its predicate dependency graph is in \cref{fig:predicate_dependencies}.
  Because of the negation in \cref{eq:example_clause} (as seen in
  \cref{fig:example_tree2}), the edge from $\mathsf{mother}$ to
  $\mathsf{father}$ is negative, while the other two edges are positive.

  We can now list the dependencies of each predicate:
  \begin{alignat*}{3}
    D_{\mathsf{parent}} &= \{ \mathsf{parent} \}, \quad && D_{\mathsf{sibling}}
    &&= \{\mathsf{sibling}, \mathsf{parent} \},\\
    D_{\mathsf{mother}} &= \{ \mathsf{mother} \}, \quad && D_{\mathsf{father}}
    &&= \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent} \}.
  \end{alignat*}
  Hence, we have two pairs of independent predicates, i.e., $\mathsf{mother}$ is
  independent from $\mathsf{parent}$ and $\mathsf{sibling}$.
\end{example}

\begin{definition}[Adjacency matrix representation]
  An $|\predicates{}| \times |\predicates{}|$ adjacency matrix $\mathbf{A}$ with
  $\{ 0, 1 \}$ as its domain is defined by
  \begin{align*}
    \mathbf{A}[i][j] = 0 \iff &\nexists k \in \{ 0, \dots, \maxNumClauses{} \}: \variable{headsOfClauses}[k].\variable{predicate} = j \text{ and } \\
    &i \in \{ a.\variable{name} \mid a \in \variable{bodiesOfClauses}[k].\variable{treeValues} \}.
  \end{align*}
\end{definition}

% TODO: continue writing from here
We define a \emph{dependency} as an algebraic data type that is either
determined (in which case it holds only the index of the \texttt{predicate}) or
undetermined (in which case it also holds the indices of the \texttt{source} and
\texttt{target} vertices, corresponding to the edge responsible for making the
dependency undetermined (exactly one, otherwise we don't consider the edge)).

Propagation for independence:
\begin{itemize}
\item Look up the dependencies of both predicates. For each pair of
  matching dependencies:
  \begin{itemize}
  \item If both are determined, fail.
  \item If one is determined, the selected edge of the other must not
    exist.
  \end{itemize}
\end{itemize}

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$} \times
    \getDependencies{$p_2$}$ such that $d_1.\predicate = d_2.\predicate$}{
    \lIf{$d_1$.\isDetermined{} {\bf and} $d_2$.\isDetermined{}}{\fail{}}
    \uIf{$d_1$.\isDetermined{}}{
      $\mathbf{A}[d_2.\source][d_2.\target]$.\removeValue{$1$}\;
    }
    \ElseIf{$d_2$.\isDetermined{}}{
      $\mathbf{A}[d_1.\source][d_1.\target]$.\removeValue{$1$}\;
    }
  }
  \caption{Propagation}
\end{algorithm}

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$} \times
  \getDependencies{$p_2$} \mid d_1.\predicate = d_2.\predicate \}$\;
  \If{$\exists (d_1, d_2) \in D$ such that $d_1.\isDetermined{}$ {\bf and}
    $d_2.\isDetermined{}$}{
    \Return{\textsc{false}}\;
  }
  \lIf{$D = \emptyset$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment}
\end{algorithm}

\begin{algorithm}[p]
  \SetKwData{edgeExists}{edgeExists}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{an $n \times n$ adjacency matrix $\mathbf{A}$}
  \Fn{\getDependencies{$p$}} {
    $D \gets \{ p \}$\;
    \While{{\normalfont \texttt{true}}}{
      $D' \gets \emptyset$\;
      \For{$d \in D$}{
        \For{$i \gets 1$ \KwTo $n$}{
          $\edgeExists \gets \mathbf{A}[i][d.\predicate] = \{ 1 \}$\;
          \uIf{$\edgeExists$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ i \}$\;
          }
          \uElseIf{$\edgeExists$ {\bf and not} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, d.\source, d.\target) \}$\;
          }
          \ElseIf{$|\mathbf{A}[i][d.\predicate]| > 1$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, i, d.\predicate) \}$\;
          }
        }
      }
      \lIf{$D' = D$}{\Return{$D$}}
      $D \gets D'$\;
    }
  }
  \caption{Computing the dependencies of a predicate}
\end{algorithm}

\section{Entailment Checking for Negative/All Cycles}
% TODO: explain that the hasNegativeCycles algorithm is an extension of a simple
% cycle detection algorithm.

\begin{algorithm}[t]
  \KwData{a constraint model for a logic program $\mathscr{P}$ amid execution}
  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
  $\mathscr{P}$ with all bodies and all predicates in heads fully
  determined\footnotemark\;
  \lIf{$\mathscr{R} = \emptyset$}{\Return{\textsc{undefined}}}
  \lIf{\hasNegativeCycles($G_{\mathscr{R}}$)}{\Return{\textsc{false}}}
  \lIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment checking for negative cycles}
\end{algorithm}
\footnotetext{Due to the definition of independence, the arguments of the head
  atom need not be determined.}
